---
title: "Analysis Report"
author: "Maria Lamine"
date: "2026/07/27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# 1. Data Importing and Cleaning Pipeline

```{r import-and-clean}
library(tidyverse)
library(janitor)
library(knitr) 

# 1. Load Raw Files
meta_raw <- read.csv("data/raw/Meta Data FINAL v2.csv")
m1_raw   <- read.csv("data/raw/measurements_part ONE_old.csv")
m2_raw   <- read.csv("data/raw/Results_2_NEW (use this).csv")

# 2. CLEAN METADATA
meta_clean <- meta_raw %>%
  clean_names() %>%
  mutate(across(everything(), str_trim)) 

# 3. CLEAN MEASUREMENTS
m1_clean <- m1_raw %>% 
  clean_names() %>% 
  mutate(across(everything(), str_trim))
  
m2_clean <- m2_raw %>% 
  clean_names() %>% 
  mutate(across(everything(), str_trim)) %>%
  rename(sample = sample_name)

# 4. Combine raw measurements
measurements_combined <- bind_rows(m1_clean, m2_clean)

# 5. Export intermediate files
if (!dir.exists("data/cleaned")) dir.create("data/cleaned", recursive = TRUE)
write_csv(meta_clean, "data/cleaned/metadata_clean.csv")
write_csv(measurements_combined, "data/cleaned/measurements_combined_clean.csv")

```

# 2. Sample Overlap Analysis

```{r overlap-summary}
# 1. Gather all unique sample IDs across metadata and combined measurements
all_samples <- sort(unique(c(meta_clean$sample_id, measurements_combined$sample)))

# 2. Build the overlap logic matrix
overlap_table <- tibble(sample_id = all_samples) %>%
  mutate(
    in_meta     = sample_id %in% meta_clean$sample_id,
    in_part1    = sample_id %in% m1_clean$sample,
    in_part2    = sample_id %in% m2_clean$sample,
    in_combined = sample_id %in% measurements_combined$sample
  )

# 3. Create a high-level summary count
overlap_summary <- overlap_table %>%
  summarise(
    total_unique_samples = n(),
    in_both_meta_and_measurements = sum(in_meta & in_combined),
    in_meta_only = sum(in_meta & !in_combined),
    in_measurements_only = sum(!in_meta & in_combined)
  )

# 4. Display summary table formatted for the report
kable(overlap_summary, caption = "Sample Overlap Summary Across Datasets")
# 4. Display summary table formatted for the report
kable(overlap_summary, caption = "Sample Overlap Summary Across Datasets")
```

# 3. Metadata Analysis

```{r metadata-analysis}
# 1. Check for missing or blank values across metadata columns
meta_missing_summary <- meta_clean %>%
  summarise(across(everything(), ~ sum(is.na(.) | . == "" | . == "NA"))) %>%
  pivot_longer(
    cols = everything(), 
    names_to = "Column Name", 
    values_to = "Missing Count"
  )

# 2. Summarize group distributions (Treatment Group x Genotype x Tissue)
meta_group_summary <- meta_clean %>%
  group_by(treatment_group, geno_type, tissue_type) %>%
  summarise(
    Total_Samples = n(),
    .groups = "drop"
  )

# 3. Display tables in the report
kable(meta_missing_summary, caption = "Missing Value Summary in Metadata")
kable(meta_group_summary, caption = "Sample Counts by Group and Tissue Type")
```

# 4. Duplicate & Conflicting Measurement Resolution

```{r resolve-duplicates}
# 1. Convert gene measurements to numeric values
measurements_num <- measurements_combined %>%
  mutate(across(c(gene_a, gene_b, gene_c), as.numeric))

# 2. Identify duplicate sample entries
duplicate_summary <- measurements_num %>%
  group_by(sample) %>%
  filter(n() > 1) %>%
  summarise(
    duplicate_count = n(),
    .groups = "drop"
  )

# 3. Resolve duplicates by averaging numerical values per sample ID
measurements_clean <- measurements_num %>%
  group_by(sample) %>%
  summarise(
    across(c(gene_a, gene_b, gene_c), ~ mean(., na.rm = TRUE)),
    .groups = "drop"
  )

# 4. Display tables in the report
kable(duplicate_summary, caption = "Duplicate Samples Identified")
kable(head(measurements_clean), caption = "Cleaned Measurements (First 6 Rows)")
```

------------------------------------------------------------------------

# 5. Quality Control (QC) Visualizations

```{r qc-plots, fig.width=8, fig.height=5}
  # Merge cleaned metadata with measurements for QC plotting
qc_data <- inner_join(meta_clean, measurements_clean, by = c("sample_id" = "sample")) %>%
  mutate(mean_value = rowMeans(pick(gene_a, gene_b, gene_c), na.rm = TRUE))
  # Check if there's a categorical factor to plot against
  group_var <- names(meta_clean)[names(meta_clean) != "sample_id"][1]
  if (!is.na(group_var)) {
    ggplot(qc_data, aes(x = .data[[group_var]], y = mean_value, fill = .data[[group_var]])) +
      geom_boxplot(alpha = 0.7, outlier.color = "red") +
      geom_jitter(width = 0.2, alpha = 0.5) +
      theme_minimal() +
      labs(
        title = paste("Quality Control: Measurement Distribution across", group_var),
        x = str_to_title(group_var),
        y = "Measurement Value"
      ) +
      theme(legend.position = "none")
  }
```
# 5. Export cleaned data
write.csv(meta_clean, "data/cleaned/meta_metadata.csv", row.names = FALSE)
write.csv(measurements_combined, "data/cleaned/measurements_combined.csv", row.names = FALSE)
# 2. CLEAN METADATA (updated)
meta_clean <- meta_raw %>%
  clean_names() %>%
  mutate(across(everything(), str_trim)) %>%
  mutate(sex = case_when(
    tolower(sex) %in% c("f", "female") ~ "F",
    tolower(sex) %in% c("m", "male") ~ "M",
    TRUE ~ sex # Leaves any blank or missing values alone
  ))
# 2. CLEAN METADATA (updated)
meta_clean <- meta_raw %>%
  clean_names() %>%
  mutate(across(everything(), str_trim)) %>%
  mutate(sex = case_when(
    tolower(sex) %in% c("f", "female") ~ "F",
    tolower(sex) %in% c("m", "male") ~ "M",
    TRUE ~ sex # Leaves any blank or missing values alone
  ))
  